2013 Western States Endurance Run Lottery
=========================================
Last update by Benjamin Chan (<benjamin.ks.chan@gmail.com>) on `r paste(Sys.time())` using `r R.version.string`.

I was a little intrigued by how the Western States Endurance Run (WSER) calculated their lottery odds estimates. Their estimates can be found in the WSER 2012 [lottery details](http://www.wser.org/2012/12/03/dec-8-lottery-details).

On 12/5/2012, WSER updated their lottery details from the version they originally posted on 12/3/2012 ([cached version](http://webcache.googleusercontent.com/search?q=cache:x6E-Gb-wm_wJ:www.wser.org/2012/12/03/dec-8-lottery-details/+&cd=2&hl=en&ct=clnk&gl=us)). The revised probabilities more closely match the results I was coming up with. The number of applicants and the number of tickets also have been revised from what was on the website on 12/3/2012.

> Updated 12/5
> 
> As posted on the [lottery applicant](http://www.wser.org/lottery2013.html) page, we have 2295 total lottery applicants for the 2013 race.
> 
> 122 applicants with four tickets = 488 tickets  
> 207 applicants with three tickets = 621 tickets  
> 480 applicants with two tickets = 960 tickets  
> 1486 applicants with one ticket = 1486 tickets
> 
> Total tickets in the hat = 3555
> 
> Here is the list of tickets ([178 page pdf](http://www.wser.org/wp-content/uploads/2012/12/2013LotteryTickets.pdf))
> 
> We are going to draw 270 unique names in the lottery and then an additional five from the lottery within the lottery. That is, the folks in the audience who have not been selected up to that point. Based on 3555 total tickets and 270 names drawn, the odds of getting selected have been updated as follows:
> 
> one ticket odds = 7.9%  
> two ticket odds = 15.2%  
> three ticket odds = 21.9%  
> four ticket odds = 28.0%
> 
> We used a Monte Carlo simulation to calculate these updated odds.

What are labeled as *odds* are really *probabilities*.

Here, I run a simulation of the lottery process to estimate probabilities of winning a slot for the Western States Endurance Run. The simulation does a few things
* Accounts for the changing probability distribution of the lottery hat as runners are selected
* Once a runner is selected, their tickets are withdrawn from the hat
* For each simulated lottery, the distribution of selected 1-ticket, 2-ticket, 3-ticket, and 4-ticket runners is determined
* From this distributions of the simulated lotteries, the estimated selection probabilities are characterized
* Selection probability distributions are plotted
* Selection distributions from the simulated lotteries are also plotted

Load the required packages.
```{r LoadPackages}
require(xtable, quietly=TRUE)
require(ggplot2, quietly=TRUE)
```


Setting up initial conditions
-----------------------------
Here is the code to set up the lottery hat data frame at the initial state. Print out some validation output just to make sure the initial state is set up correctly.
```{r Setup}
spots <- 270
distn <- c(1486, 480, 207, 122)
applicants <- sum(distn)
runner <- seq(1, applicants)
tickets <- c(rep(4, distn[4]), rep(3, distn[3]), rep(2, distn[2]), rep(1, distn[1]))
dfHat <- data.frame(runner, tickets)
dfHat$prob <- dfHat$tickets / sum(dfHat$tickets)
table(factor(dfHat$tickets))
head(dfHat)
tail(dfHat)
```


Lottery simulation
------------------
The simulation needs to account for the changing relative distribution of tickets after a person is selected and their tickets are no longer in the pool of eligible tickets.

The matrix `lottery` is an $I \times J$ matrix where row $i$ is the $i$-th simulation and the column $j$ is the $j$-th lottery winner drawn. The number of columns in the matrix is `r spots`, variable `spots`. The number of simulated lotteries is variable `size`. Set the random number seed as the date of the lottery in numeric form multipied by the number of applicants.
```{r Simulation}
size <- 1E4
set.seed(as.numeric(as.Date("2012-12-08", format="%Y-%m-%d")) * applicants)
lottery <- matrix(nrow=size, ncol=spots)
start <- Sys.time()
for (i in 1:size) {
  lottery[i, ] <- sample(dfHat$runner, spots, prob=dfHat$prob)
}
end <- Sys.time()
end - start
```

Here's an example of the selected runners drawn from a random simulated lottery.
```{r RandomLotteryResult}
i <- sample(seq(1, size), 1)
sampLottery <- list(i, sort(lottery[i, ]))
names(sampLottery) <- c("lottery", "runner")
sampLottery
```
Here's the distribution of the category of ticket holders from that random simulated lottery.
```{r RandomLotteryDistn}
table(dfHat$tickets[sampLottery$runner])
```
I.e., in simulated lottery `r i`, 
* `r sum(dfHat$tickets[sampLottery$runner] == 1)` applicants with 1 ticket were selected  (`r format(100 * sum(dfHat$tickets[sampLottery$runner] == 1) / distn[1], digits=2)`%)
* `r sum(dfHat$tickets[sampLottery$runner] == 2)` applicants with 2 tickets were selected (`r format(100 * sum(dfHat$tickets[sampLottery$runner] == 2) / distn[2], digits=2)`%)
* `r sum(dfHat$tickets[sampLottery$runner] == 3)` applicants with 3 tickets were selected (`r format(100 * sum(dfHat$tickets[sampLottery$runner] == 3) / distn[3], digits=2)`%)
* `r sum(dfHat$tickets[sampLottery$runner] == 4)` applicants with 4 tickets were selected (`r format(100 * sum(dfHat$tickets[sampLottery$runner] == 4) / distn[4], digits=2)`%)

Okay... but what happened with the other `r size - 1` simulated lotteries?

Format lottery simulation data
------------------------------
I'm not really interested in which runners were selected in the lottery simulation. What I'm really after are estimates for the probability of selecting a runner, among the `r spots` available spots, with $X$ tickets in the initial hat.

To get at this, first I'll have to match the runners selected to the number of tickets they started out with.
```{r MatchTickets}
lottery2 <- matrix(nrow=size, ncol=spots)
for (i in 1:size) {
  lottery2[i, ] <- dfHat$tickets[lottery[i, ]]
}
```
Reformat the `lottery2` matrix to an aggregated data frame for analysis.
```{r ReformatMatrix}
start <- Sys.time()
tickets <- factor(as.vector(t(lottery2)))
sim <- rep(seq(1, size), each=spots)
dfLottery <- data.frame(sim, tickets)
aggLottery <- aggregate(tickets ~ sim, dfLottery, table)
sim <- rep(seq(1, size), each=4)
tickets <- factor(rep(seq(1, 4), size))
freq <- as.vector(t(aggLottery$tickets))
dfSummary <- data.frame(sim, tickets, freq)
end <- Sys.time()
end - start
```
Save the aggregated data frame for other analysis.
```{r Save}
setwd("~/Dropbox/Sandbox/WSERLottery")
save(aggLottery, file="aggLottery.RData")
```

For each type of lottery applicant (1 ticket, 2 tickets, etc.), calculate the proportion of selected applicants. 
```{r Probabilities}
total <- rep(distn, size)
dfSummary$prob <- 100 * (dfSummary$freq / total)
avg <- aggregate(prob ~ tickets, dfSummary, mean)
med <- aggregate(prob ~ tickets, dfSummary, median)
sd  <- aggregate(prob ~ tickets, dfSummary, sd)
ev <- distn * avg[, 2] / 100
probWSER <- c(7.9, 15.2, 21.9, 28.0)
evWSER <- distn * probWSER / 100
diffprob <- avg[, 2] - probWSER
diffev <- ev - evWSER
pctdiff <- 100 * diffprob / avg[, 2]
simsum <- data.frame(avg, med[, 2], sd[, 2], distn, ev, probWSER, evWSER, diffprob, diffev, pctdiff)
names(simsum) <- c("Tickets", "Mean", "Median", "SD", "N", "EV", "Prob (WSER)", "EV (WSER)", "Diff. prob.", "Diff. EV", "% diff.")
```

Summarize lottery simulations
-----------------------------
Plot the distribution of probabilities from the `r format(size, big.mark=",")` simulated lotteries. Annotate with the estimated mean selection probability.
```{r PlotProbabilities}
y1 <- max(density(dfSummary$prob[dfSummary$tickets == 1])$y)
y2 <- max(density(dfSummary$prob[dfSummary$tickets == 2])$y)
y3 <- max(density(dfSummary$prob[dfSummary$tickets == 3])$y)
y4 <- max(density(dfSummary$prob[dfSummary$tickets == 4])$y)
y <- c(y1, y2, y3, y4)
ggplot(dfSummary, aes(x=prob, y=..density.., fill=tickets)) +
  geom_density(alpha=1/2, color=NA) +
  scale_fill_brewer(type="div", palette="BrBG") +
  labs(title="2012 WSER Lottery Selection Probability Densities", x="Percent", y=paste("Proportion of", format(size, big.mark=","), "simulations"), fill="Tickets") +
  annotate("text", label=paste(format(simsum$Mean, digits=2, trim=TRUE), "%", sep=""), x=simsum$Mean, y=y) +
  theme(legend.position="bottom")
```
As expected, the spread of the selection probabilities increases as the number of tickets a person has in the hat increases (the variance of a binomial random variable increases with $p$).

Another way to think about the lottery is to plot the distribution of the frequency of runners selected by number of tickets. Annotate with the estimated expected value.
```{r PlotFrequencies}
y1 <- max(density(dfSummary$freq[dfSummary$tickets == 1])$y)
y2 <- max(density(dfSummary$freq[dfSummary$tickets == 2])$y)
y3 <- max(density(dfSummary$freq[dfSummary$tickets == 3])$y)
y4 <- max(density(dfSummary$freq[dfSummary$tickets == 4])$y)
y <- c(y1, y2, y3, y4)
ggplot(dfSummary, aes(x=freq, y=..density.., fill=tickets)) +
  geom_density(alpha=1/2, color=NA) +
  scale_fill_brewer(type="div", palette="BrBG") +
  labs(title="2012 WSER Lottery Selection Distribution Densities", x="Number", y=paste("Proportion of", format(size, big.mark=","), "simulations"), fill="Tickets") +
  annotate("text", label=format(simsum$EV, digits=2, trim=TRUE), x=simsum$EV, y=y) +
  theme(legend.position="bottom")
```

Print a table summarizing the simulated lotteries showing the mean and median selection probabilities and their standard deviations. Compare to probabilities given in the WSER 2012 [lottery details](http://www.wser.org/2012/12/03/dec-8-lottery-details).
```{r Table, results='asis'}
print(xtable(simsum), type="html", include.rownames=FALSE)
```
My estimates and the probabilities calculated by WSER are essentially identical. Percent difference is never more than `r max(abs(simsum$"% diff."))`%.

Plot the outcomes of a random sample of the `r format(size, big.mark=",")` simulated lotteries as a [waffle plot](http://www.improving-visualisation.org/vis/id=179). The width of each bar represents the number of selected runners. Blocks represent 10 runners.
```{r PlotLotteryResults, fig.height=15}
s <- 100
i <- sample(seq(1, size), s)
dfSample <- dfLottery[dfLottery$sim %in% i, ]
dfSample$sim <- factor(dfSample$sim)
levels(dfSample$sim) <- rev(levels(dfSample$sim))
ggplot(dfSample, aes(x=sim, fill=tickets)) +
  geom_bar(width=1) +
  geom_hline(y=seq(0, spots, 10), color="white") +
  geom_vline(x=seq(1, s)-0.5, color="white") +
  scale_fill_brewer(type="div", palette="BrBG") +
  scale_y_continuous(expand=c(0, 0)) +
  labs(title=paste("Simulated 2012 WSER Lotteries\n", "Sample of", s, "Lotteries"), x="Simulated lottery", y="Number of selected runners", fill="Tickets") +
  coord_flip() +
  theme(legend.position="top")
```


Actual results
--------------
The actual lottery was held on December 8, 2012. Here are the results.

> Percentages for tickets holders drawn in lottery today:  
> 1 Ticket - 128 8.6%  
> 2 Ticket - 68 14.1%  
> 3 Ticket - 37 17.8%  
> 4 Ticket - 42 34.4%  

These results include the 5 winners of the 
> Bonus Drawing at the end of the Lottery giv[ing] those present “one last chance” to be selected.

So I'll need to subtract out the 5 winners of the bonus drawing. Three of the bonus drawing winners were 1-ticket holders, 1 was a 3-ticket holder, and 1 was a 4-ticket holder.

Compare the observed lottery to the expected lottery.
```{r ObsCompare, results='asis'}
yObs <- c(128 - 3, 68, 37 - 1, 42 - 1)
yExp <- simsum$EV
dfObsExp <- data.frame(yObs, yExp, diff=yObs - yExp)
names(dfObsExp) <- c("Observed", "Expected", "Difference")
print(xtable(dfObsExp), type="html", include.rownames=FALSE)
```

I wonder how many times this outcome appeared in my simulated lotteries. 
```{r SimObserved}
y1 <- aggLottery$tickets[,1] == yObs[1]
y2 <- aggLottery$tickets[,2] == yObs[2]
y3 <- aggLottery$tickets[,3] == yObs[3]
y4 <- aggLottery$tickets[,4] == yObs[4]
y <- y1 & y2 & y3 & y4
aggLottery[y,]
```
Of the `r format(size, big.mark=",")` simulated lotteries, only `r sum(y)`, or `r 100 * sum(y) / size`%, matched the exact outcome of the actual lottery.
